### 2025-06-18 chapters/background.tex
reorganised key sections
updated sections 1.3, 2.3 and included more citations

\chapter{Background and System Architecture}
\label{chap:background}

@@

Overall, Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization. Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}.

\section{Comparison with Other TEEs}

@@

>>>>

\chapter{Background}
\label{chap:background}

@@

Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization. Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}. 
To support secure communication and attestation, Keystone enclaves can integrate modern cryptographic primitives, including emerging post-quantum schemes such as Kyber.

\section{Kyber: A Post-Quantum Cryptography Algorithm}
\label{sec:kyber}

Kyber is a post-quantum key encapsulation mechanism (KEM) designed to resist attacks from both classical and quantum adversaries. Its security is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem considered intractable even for quantum computers. Kyber is one of the finalists in the NIST Post-Quantum Cryptography (PQC) competition and has been selected for standardization due to its strong security and performance characteristics \cite{kyber2024}.  % [SOURCE1]

At its core, Kyber is derived from the Lyubashevsky–Peikert–Regev (LPR) encryption framework, which adapts Learning With Errors (LWE) from integer vectors to more efficient polynomial structures. Specifically, Kyber operates in the Module-LWE setting, which balances the structure of Ring-LWE with the flexibility of traditional LWE. This results in strong performance and compact key sizes while preserving security \cite{kyber2024,kyber2021}. % [SOURCE1, BLOG1]

Kyber supports three security levels:
\begin{itemize}
    \item \textbf{Kyber-512}, targeting AES-128-level security,
    \item \textbf{Kyber-768}, targeting AES-192-level security (recommended),
    \item \textbf{Kyber-1024}, targeting AES-256-level security.
\end{itemize}

\begin{table}[h]
\centering
\caption{Kyber and RSA: Key and Ciphertext Sizes (in bytes)}
\label{tab:kyber_sizes}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
\hline
Kyber-512 & 1632 & 800 & 768 \\
Kyber-768 & 2400 & 1184 & 1088 \\
Kyber-1024 & 3168 & 1568 & 1568 \\
RSA-3072 & 384 & 384 & 384 \\
RSA-15360 & 1920 & 1920 & 1920 \\
\hline
\end{tabular}
\end{table}

While RSA keys are somewhat smaller, Kyber maintains significantly better security in the post-quantum context, and its keys are far more compact than those of many alternative PQC candidates such as Classic McEliece \cite{kyber2021}. % [BLOG1]

Kyber is structured as a KEM rather than a traditional public-key encryption (PKE) scheme. Internally, however, it uses a PKE system transformed into a CCA-secure KEM. The KEM exposes a simple API:
\begin{itemize}
    \item $(\mathit{pk}, \mathit{sk}) = \texttt{KeyGen}()$
    \item $(\mathit{ct}, \mathit{ss}) = \texttt{Encapsulate}(\mathit{pk})$
    \item $\mathit{ss} = \texttt{Decapsulate}(\mathit{sk}, \mathit{ct})$
\end{itemize}

The encapsulation mechanism makes Kyber especially suitable for protocols like TLS, Signal, and OpenPGP. % [BLOG1]

A simplified pedagogical version called ``Baby Kyber’’ is often used to illustrate the cryptographic structure. Baby Kyber uses polynomials over a small modulus and limited degree to show how private and public keys, as well as ciphertexts, are constructed and manipulated. Encryption relies on adding small random errors to hide the message, while decryption relies on error tolerance due to coefficient scaling. If the message polynomial is scaled appropriately, decryption can recover the original bit string by rounding \cite{kyber2021}. % [BLOG1]

In full Kyber, the parameters are chosen to ensure security and correctness:
\begin{itemize}
    \item $n = 256$: polynomial degree
    \item $k \in \{2, 3, 4\}$: number of polynomials (Kyber-512, 768, 1024)
    \item $q = 3329$: prime modulus
    \item $\eta_1, \eta_2$: bounds on error distributions
    \item $d_u, d_v$: compression factors
\end{itemize}

Kyber’s performance has been optimized for a variety of platforms. Table~\ref{tab:kyber_perf} shows benchmark results on an Intel Haswell CPU with AVX2 vectorization \cite{kyber2024}. % [SOURCE1]

\begin{table}[h]
\centering
\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
\label{tab:kyber_perf}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Variant} & \textbf{KeyGen (cycles)} & \textbf{Encaps (cycles)} & \textbf{Decaps (cycles)} \\
\hline
Kyber-512 & 33,856 & 45,200 & 34,572 \\
Kyber-768 & 52,732 & 67,624 & 53,156 \\
Kyber-1024 & 73,544 & 97,324 & 79,128 \\
\hline
\end{tabular}
\end{table}

Kyber also includes a variant called \texttt{Kyber-90s}, which replaces the use of SHAKE with AES-256 and SHA2 primitives, enabling further acceleration on platforms with AES instruction support \cite{kyber2024}. % [SOURCE1]

The security of Kyber ultimately reduces to the hardness of the Shortest Vector Problem (SVP) via the Module-LWE problem. Recovering a message or key requires solving an instance of SVP in a structured lattice, a task that remains intractable even with quantum resources\cite{kyber2021}. % [BLOG1]

Due to its efficient structure and compact representation, Kyber has been adopted by numerous organizations:
\begin{itemize}
    \item Cloudflare included Kyber in the CIRCL cryptographic library
    \item Amazon integrated hybrid Kyber modes into AWS KMS
    \item IBM deployed Kyber in its quantum-secure tape drive
\end{itemize} % [SOURCE1]
These deployments demonstrate Kyber’s readiness for practical applications in secure communications, data protection, and cryptographic protocols. 
The integration of Kyber into TEE frameworks like Keystone is increasingly relevant as systems must remain secure even in the presence of quantum-capable adversaries. TEEs benefit from Kyber’s compact keys and efficient operations, making it a strong candidate for enclave-based key exchange and attestation.

\section{Comparison with Other TEEs}
\label{sec:comparison_tees}

@@

%\section{Post-Quantum Cryptography and Kyber}
%\section{Performance Considerations for PQC in TEEs}

### 2025-06-18 chapters/introduction.tex
updated sections 1.3, 2.3 and included more citations

This thesis is organized as follows: Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs). Chapter~\ref{chap:background} presents the technical foundations and system architecture of the Keystone framework, including a comparative analysis with Intel SGX and ARM TrustZone that highlights key design trade-offs and performance considerations. Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation. Chapter~\ref{chap:results} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights. Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements. Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.

>>>>

This thesis is organized as follows: 
\\
Chapter~\ref{chap:intro} introduces the motivation for secure computation in untrusted environments, outlines the core contributions of this work, and surveys relevant literature in the domain of Trusted Execution Environments (TEEs).

Chapter~\ref{chap:background} provides foundational background on Trusted Execution Environments (TEEs), with a particular emphasis on the Keystone framework and the RISC-V architecture. It also introduces Kyber, a post-quantum key encapsulation mechanism, and analyzes how its characteristics align with secure enclave execution.

Chapter~\ref{chap:methodology} describes the experimental setup, detailing the virtualized testbed, benchmark selection, parameter tuning, and data collection methods employed to ensure rigorous and reproducible evaluation.

Chapter~\ref{chap:benchmarking} reports empirical findings and analyzes the impact of system parameters such as CPU core count, memory allocation, and cache size on enclave performance, along with associated overheads and insights.

%Chapter~\ref{chap:recommendations} synthesizes these results into practical configuration guidelines for deploying Keystone TEEs across diverse workload scenarios and identifies architectural bottlenecks for future improvements.%

Chapter~\ref{chap:discussion} discusses the trade-offs revealed by the experiments and considers the implications for future secure computing architectures.

Finally, Chapter~\ref{chap:conclusion} summarizes the main contributions and discusses their implications for secure system design, outlining promising directions for further research including hardware extensions and formal verification.

### 2025-06-18 chapters/system-architecture.tex
reorganised key sections
updated sections 1.3, 2.3 and included more citations

\chapter{Methodology}
\label{chap:methodology}
The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads. Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment. To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.

@@

\section{Benchmarking tools and metrics}

@@

The Dhrystone benchmark has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern. Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications. Because of its simplicity and minimal memory usage, Dhrystone provides a good baseline measure of a processor’s raw computing capability. For this study, the RISC-V version of Dhrystone was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware. The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave. This setup made it possible to measure the specific performance overhead introduced by Keystone’s security features, such as memory isolation and the cost of switching between secure and non-secure execution contexts.

@@

Both Dhrystone and CoreMark are CPU-intensive, which is exactly what makes them useful in this context. Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance. CoreMark, on the other hand, introduces a bit more complexity and memory interaction, which makes it better suited for revealing the broader impact of Keystone’s security mechanisms. Neither benchmark uses file I/O or persistent storage, which means that this study does not explore storage-related features of the TEE, such as secure disk access or file encryption. That’s outside the scope here and would require a different kind of benchmarking approach.

>>>>

\chapter{System Architecture and Design}
\label{chap:methodology}
The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads. Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment. To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.

@@

\section{Kyber in the Enclave}
\label{sec:kyber-enclave}

\section{Benchmarking tools and metrics}
\label{sec:benchmarking-tools}

@@

The Dhrystone benchmark\cite{weiss2002dhrystone} has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern. Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications. Because of its simplicity and minimal memory usage, Dhrystone\cite{weiss2002dhrystone} provides a good baseline measure of a processor’s raw computing capability. For this study, the RISC-V version of Dhrystone\cite{weiss2002dhrystone} was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware. The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave. This setup made it possible to measure the specific performance overhead introduced by Keystone’s security features, such as memory isolation and the cost of switching between secure and non-secure execution contexts.

@@

Both Dhrystone\cite{weiss2002dhrystone} and CoreMark\cite{gal2012exploring} are CPU-intensive, which is exactly what makes them useful in this context. Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance. CoreMark, on the other hand, introduces a bit more complexity and memory interaction, which makes it better suited for revealing the broader impact of Keystone’s security mechanisms. Neither benchmark uses file I/O or persistent storage, which means that this study does not explore storage-related features of the TEE, such as secure disk access or file encryption. That’s outside the scope here and would require a different kind of benchmarking approach.
